---
header-includes:
- \usepackage{xcolor}
- \usepackage{color} 
- \usepackage{fancyhdr,color}
- \usepackage{lipsum}
- \fancyfoot[CE] {\thepage}
title: "**Final Project AFCS \n &nbsp;  \n Team 1**"
subtitle: "Applied Forecasting in Complex Systems 2024"
author: Alberto Scinetti (13115316), Rosa Keuss (12690589), Ngoc Doan (11576057)
Simon Hüsgen (15215555), and Pearl Owusu (12502340)
date: "University of Amsterdam \n &nbsp;  \n December, 20, 2024 "
output: pdf_document
fontsize: 11pt
highlight: tango
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  dev.args = list(pointsize = 11)
)
options(digits = 3, width = 60)
library(fpp3)
library(latex2exp)
library(tidyverse)
library(prophet)
library(tsibble)
```

## Data Description / Exploration

The data set "sales" ("sales_train_and_validation") includes 823 products from the "FOODS_3" category (assumed to be category Food and Department 3) sold by a Walmart store uniquely identified as "TX3" in Texas (US) for the subsequent 1912 days ( $\approx$ 5 years and 3 months) starting on 2011-01-29. Each product within this category is uniquely identified by a code, such as "FOODS_3_007".

The "calendar" data set contains the information about the dates the product are sold, information that might be useful for modelling future sales based on weekday/weekend/event day.

The "sell_prices" data set contains information about the price of the products sold per store and date (also might be useful to include this in the model but maybe less important than date? not sure)

Other data sets are for testing and evaluation.

*Importing the data sets:*

```{r}
calendar <- read.csv("calendar_afcs2024.csv")
sell_prices <- read.csv("sell_prices_afcs2024.csv")
sales_train_and_validation <- read.csv("sales_train_validation_afcs2024.csv")
sales_test_and_validation <- read.csv("sales_test_validation_afcs2024.csv") 
#sample_submission <- read.csv("sample_submission_afcs2024.csv")
sales_test_and_evaluation <- read.csv("sales_test_evaluation_afcs_2024.csv") 

# How many products to make forecast on? 823 
print(nrow(sales_train_and_validation))
print(nrow(sales_test_and_validation))
```

*Convert main sales data to a tsibble object and visualize number of stock sold per day for a product.*

In this case we visualize number of products sold over the time span for product FOODS_3_001 and FOODS_3_007.

Note that different products have different fluctuation, different distribution and different peaks (like product 007 is sold in such a higher quantity than product 001)

```{r}
# we convert the prediction object into a tsibble 

df_long_sales <- sales_train_and_validation %>%
  pivot_longer(cols = starts_with("d_"), 
               names_to = "day", 
               values_to = "items_sold") %>%
  mutate(day = as.integer(sub("d_", "", day))) 


df_long_sales <- df_long_sales %>%
  mutate(date = as.Date("2011-01-29") + days(day - 1))  

df_sales_tsibble <- df_long_sales %>%
  as_tsibble(index = date, key = id)

```

```{r}
# also add calendar features to the training data (day of the week, month, holiday and snap )
calendar <- calendar %>%
  mutate(date = as.Date(date, format = "%m/%d/%Y"))

# holiday indicator 
calendar <- calendar  %>%
  mutate(
    is_holiday = ifelse(!is.na(event_name_1) | !is.na(event_name_2), 1, 0),
    holiday_type = coalesce(event_type_1, event_type_2)  # Combine types into one column
  )

# merge 
df_merged <- df_long_sales %>%
  left_join(calendar %>% select(date, weekday, month, is_holiday, holiday_type, snap_TX), by = "date")

df_merged_tsibble <- df_merged %>%
  as_tsibble(index = date, key = id)

df_merged_tsibble 
```

```{r}
# visualize some data for some days
df_merged_tsibble  %>% 
  filter(id =="FOODS_3_001_TX_3_validation") %>%
  autoplot(items_sold)

df_merged_tsibble %>% 
  filter(id == "FOODS_3_090_TX_3_validation") %>%
  autoplot(items_sold) +
  labs(
    title = "Distribution of Product TX 090",
    x = "Date",
    y = "Items Sold"
  )

```

Now we explore the sell prices data set

*Need to add a visualization to see how sell prices data is missing a lot of values*

```{r}

start_date <- as.Date("2011-01-29")

# add daily sell price using weekly data
sell_prices <- sell_prices %>%
  mutate(week_start_date = start_date + weeks(wm_yr_wk - min(wm_yr_wk)))

# tsibble conv
data_sell_tsibble <- sell_prices %>%
  as_tsibble(index = week_start_date, key = c(store_id, item_id))

data_sell_tsibble <- data_sell_tsibble %>%
  mutate(item_id = paste0(item_id, "_TX_3_validation"))




# to daily data 
data_sell_daily <- data_sell_tsibble %>%
  mutate(date = map(week_start_date, ~ seq.Date(.x, by = "day", length.out = 7))) %>%
  unnest(cols = c(date))

data_sell_daily 
```

```{r}
# View the resulting tsibble
data_sell_tsibble%>% 
  filter(item_id == "FOODS_3_001_TX_3_validation") %>%
  autoplot(sell_price)


data_sell_tsibble %>% 
  filter(item_id == "FOODS_3_007_TX_3_validation") %>%
  autoplot(sell_price)

```

```{r}

# List of products to analyze
products <- c("FOODS_3_001_TX_3_validation", "FOODS_3_002_TX_3_validation",  "FOODS_3_090_TX_3_validation","FOODS_3_777_TX_3_validation", "FOODS_3_500_TX_3_validation", "FOODS_3_352_TX_3_validation",  "FOODS_3_620_TX_3_validation")

# Generate a complete sequence of weekly start dates
all_weeks <- tibble(
  week_start_date = seq(
    from = min(data_sell_tsibble$week_start_date),
    to = max(data_sell_tsibble$week_start_date),
    by = "week"
  )
)

# Function to process and visualize missing weeks for each product
process_product <- function(product_id) {
  # Filter data for the specific product
  filtered_data <- data_sell_tsibble %>%
    filter(item_id == product_id)
  
  # Join the complete sequence with the filtered data
  missing_weeks_data <- all_weeks %>%
    left_join(filtered_data, by = "week_start_date") %>%
    mutate(
      status = ifelse(is.na(sell_price), "Missing", "Present"),
      item_id = product_id  # Add product identifier
    )
  
  return(missing_weeks_data)
}

# Process all products and combine the results
missing_data_combined <- bind_rows(lapply(products, process_product))

# Visualization
ggplot(missing_data_combined, aes(x = week_start_date, y = item_id, color = status)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Present" = "blue", "Missing" = "red")) +
  labs(
    title = "Missing Weeks Visualization for Multiple Products",
    x = "Week Start Date",
    y = "Product",
    color = "Week Status"
  ) +
  theme_minimal()

# Process all products and combine the results
missing_data_combined <- bind_rows(lapply(products, process_product))

# Visualization
ggplot(missing_data_combined, aes(x = week_start_date, y = item_id, color = status)) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Present" = "blue", "Missing" = "red")) +
  labs(
    title = "Missing Weeks Visualization for Multiple Products",
    x = "Week Start Date",
    y = "Product",
    color = "Week Status"
  ) +
  theme_minimal()


```

```{r}
# convert weekly sell price to daily 
data_sell_daily <- data_sell_daily %>%
  select(date, item_id, sell_price)

# merge
df_merged_tsibble <- df_merged_tsibble %>%
  left_join(data_sell_daily, by = c("date", "id" = "item_id"))

df_merged_tsibble
```

```{r}
# look at sales heatmpat 
df_sales_tsibble %>%
  mutate(month = format(date, "%b"),  
         day_of_week = weekdays(date)) %>%  
  group_by(month, day_of_week) %>% 
  summarise(total_sales = sum(items_sold, na.rm = TRUE)) %>% 
  ggplot(aes(x = month, y = day_of_week, fill = total_sales)) +
  geom_tile() +  
  labs(title = "Sales Heatmap by Day of Week and Month", 
       x = "Month", 
       y = "Day of Week", 
       fill = "Total Sales") +
  scale_fill_viridis_c() + 
  theme_minimal() 

```

more data analysis

```{r}
library(dplyr)
library(ggplot2)

# Unique events
unique_events <- calendar %>%
  select(event_name_1, event_name_2) %>%
  summarise(
    unique_event_1 = n_distinct(event_name_1, na.rm = TRUE),
    unique_event_2 = n_distinct(event_name_2, na.rm = TRUE)
  )
print(unique_events)

# Unique event types
unique_event_types <- calendar %>%
  select(event_type_1, event_type_2) %>%
  summarise(
    unique_event_type_1 = n_distinct(event_type_1, na.rm = TRUE),
    unique_event_type_2 = n_distinct(event_type_2, na.rm = TRUE)
  )
print(unique_event_types)

# Distribution of days (weekday, month, year)
weekday_distribution <- calendar %>%
  count(weekday) %>%
  arrange(desc(n))
print(weekday_distribution)

month_distribution <- calendar %>%
  count(month) %>%
  arrange(month)
print(month_distribution)

year_distribution <- calendar %>%
  count(year) %>%
  arrange(year)
print(year_distribution)

# Distribution of the event types 
event_type_1_distribution <- calendar %>%
  count(event_type_1) %>%
  arrange(desc(n))
print(event_type_1_distribution)

event_type_2_distribution <- calendar %>%
  count(event_type_2) %>%
  arrange(desc(n))
print(event_type_2_distribution)

# Event by weekday
event_by_weekday <- calendar %>%
  group_by(weekday) %>%
  summarise(
    event_1_count = sum(!is.na(event_name_1)),
    event_2_count = sum(!is.na(event_name_2)),
    .groups = "drop"
  ) %>%
  arrange(weekday)
print(event_by_weekday)

# other variables
snap_distribution <- calendar %>%
  count(snap_TX) %>%
  arrange(snap_TX)
print(snap_distribution)

# Plots
ggplot(weekday_distribution, aes(x = reorder(weekday, -n), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Distribution of Days of the Week", x = "Weekday", y = "Count") +
  theme_minimal()

ggplot(month_distribution, aes(x = as.factor(month), y = n)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  labs(title = "Distribution of Months", x = "Month", y = "Count") +
  theme_minimal()

ggplot(event_type_1_distribution, aes(x = reorder(event_type_1, -n), y = n)) +
  geom_bar(stat = "identity", fill = "lightcoral") +
  labs(title = "Distribution of Event Types (Type 1)", x = "Event Type", y = "Count") +
  theme_minimal()

ggplot(event_type_2_distribution, aes(x = reorder(event_type_2, -n), y = n)) +
  geom_bar(stat = "identity", fill = "lightyellow") +
  labs(title = "Distribution of Event Types (Type 2)", x = "Event Type", y = "Count") +
  theme_minimal()
```

STL decomposition for an item

```{r}
library(forecast)
#one product
df_foods_001 <- df_sales_tsibble %>%
  filter(id == "FOODS_3_090_TX_3_validation")

#daily
ts_foods_001 <- ts(df_foods_001$items_sold, frequency = 365)

#STL decomposition
fit_stl <- stl(ts_foods_001, s.window = "periodic")
autoplot(fit_stl) +
  labs(
    title = "STL Decomposition of Product TX 090",
    x = "Date",
    y = "Items Sold"
  )


```

Correlation between prices and sales for an item

```{r}
#Merge sales and price for an item
merged_data <- df_sales_tsibble %>%
  filter(id == "FOODS_3_001_TX_3_validation") %>%
  left_join(data_sell_tsibble %>% filter(item_id == "FOODS_3_001_TX_3_validation"),
            by = c("date" = "week_start_date"))

#Correlation sales and price: 0.0469 for 001
cor(merged_data$sell_price, merged_data$items_sold, use = "complete.obs")
```

# Forecast Analysis (on single product)

```{r}
# mutate values to use for analys as factors to be sure they're good for analysis 
df_merged_tsibble <- df_merged_tsibble %>%
  mutate(weekday = factor(weekday, levels = c("Sunday", "Monday", "Tuesday", 
                                              "Wednesday", "Thursday", "Friday", 
                                              "Saturday")))
```

```{r}
#Convert Validation Data to Long Format
validation_tsibble <- sales_test_and_evaluation %>%
  pivot_longer(
    cols = starts_with("d_"),
    names_to = "day",
    values_to = "items_sold"
  ) %>%
  mutate(
    day = as.integer(sub("d_", "", day)),
    date = as.Date("2016-04-24") + (day - 1942)  # Adjust starting date
  ) %>%
  select(-day) %>%
  as_tsibble(index = date, key = id)

```

```{r}

# Create a tsibble for the next 28 days
future_dates <- tibble(
  date = seq.Date(
    from = max(df_merged_tsibble$date) + 1,
    by = "day",
    length.out = 28
  )
) %>%
  mutate(
    weekday = factor(weekdays(date), levels = c("Sunday", "Monday", "Tuesday", 
                                                "Wednesday", "Thursday", "Friday", 
                                                "Saturday")),
    #month = factor(month(date), levels = as.character(1:12))
  )

# Add the ID to align with the model's key
future_tsibble <- future_dates %>%
  mutate(id = "FOODS_3_090_TX_3_validation") %>%
  as_tsibble(index = date, key = id)
```

```{r}
future_tsibble <- future_tsibble %>%
  left_join(calendar %>% select(date, month, is_holiday, snap_TX), by = "date")

future_tsibble <- future_tsibble %>%
  left_join(data_sell_daily, by = c("date", "id" = "item_id"))

future_tsibble 
```

```{r}
#install.packages("fable.prophet")
library(fable.prophet)
```

```{r}
# Fit ARIMA model with covariates
models <- df_merged_tsibble %>%
  filter(id == "FOODS_3_090_TX_3_validation") %>%
  model(
    ARIMA = ARIMA(items_sold),
    ARIMA1 = ARIMA(items_sold~ weekday + snap_TX + is_holiday),
    ARIMA2 = ARIMA(items_sold ~ weekday + snap_TX),
    ETS = ETS(items_sold),
    SNAIVE = SNAIVE(items_sold),
    TSLM = TSLM(items_sold ~ trend() + season() + is_holiday + snap_TX),
    PROPHET = prophet(items_sold ~ weekday + is_holiday + snap_TX)
  )

# Forecast using the future tsibble for all models
forecasts <- models %>%
  forecast(new_data = future_tsibble)

# Clip negative predictions and round to integers
forecasts <- forecasts %>%
  mutate(
    .mean = pmax(0, round(.mean))
  )

# Plot forecasts for all models
forecasts %>%
  autoplot(df_merged_tsibble %>% filter(id == "FOODS_3_090_TX_3_validation")) +
  labs(title = "Forecast Comparison", y = "Items Sold", x = "Date") +
  facet_wrap(~.model, scales = "free_y")  # Separate plot for each model


```

```{r}
# Join forecasts with validation data
results_comparison <- forecasts %>%
  as_tibble() %>%
  select(.model, id, date, forecast = .mean) %>%
  left_join(
    validation_tsibble,  # Validation data containing actual sales
    by = c("id", "date")
  )

```

```{r}
# Calculate error metrics for each model
results_comparison <- results_comparison %>%
  mutate(
    error = abs(forecast - items_sold)  # Absolute error
  )

# MAE and RMSE by model
metrics <- results_comparison %>%
  group_by(.model) %>%
  summarise(
    MAE = mean(error, na.rm = TRUE),
    RMSE = sqrt(mean(error^2, na.rm = TRUE))
  )

# Print metrics
print(metrics)

```

```{r}
# Plot forecasts vs actuals for each model
results_comparison %>%
  ggplot(aes(x = date)) +
  geom_line(aes(y = items_sold, color = "Actual"), size = 1) +
  geom_line(aes(y = forecast, color = .model), linetype = "dashed") +
  facet_wrap(~.model, scales = "free_y") +
  labs(
    title = "Forecast vs Actuals by Model",
    y = "Items Sold",
    x = "Date"
  ) +
  theme_minimal()
```

```{r}
# Plot residual diagnostics for all models
models %>%
  select(ARIMA2) %>%
  gg_tsresiduals()

```

## **A biggest Forecast sample for validation**

```{r}
sample_ids <- df_merged_tsibble %>%
  distinct(id) %>%
  slice_sample(n = 20) %>% # Adjust `n` for the number of products you want
  pull(id)
sample_ids

sampled_data <- df_merged_tsibble %>%
  filter(id %in% sample_ids)

sample_ids 

```

```{r}
# atsibble for the next 28 days
future_dates <- tibble(
  date = seq.Date(
    from = max(df_merged_tsibble$date) + 1,
    by = "day",
    length.out = 28
  )
) %>%
  mutate(
    weekday = factor(weekdays(date), levels = c("Sunday", "Monday", "Tuesday", 
                                                "Wednesday", "Thursday", "Friday", 
                                                "Saturday"))
  )

# the sampled IDs to align with the model's key
future_tsibble <- expand_grid(
  date = future_dates$date,
  id = sample_ids  
) %>%
  left_join(future_dates, by = "date") %>% 
  as_tsibble(index = date, key = id)

future_tsibble <- future_tsibble %>%
  left_join(calendar %>% select(date, month, is_holiday, snap_TX), by = "date")

future_tsibble <- future_tsibble %>%
  left_join(data_sell_daily, by = c("date", "id" = "item_id"))

future_tsibble 
```

```{r}
models_sampled <- sampled_data %>%
  model(
    ARIMA = ARIMA(items_sold),
    ARIMA1 = ARIMA(items_sold ~ weekday + snap_TX + is_holiday),
    ARIMA2 = ARIMA(items_sold ~ weekday + snap_TX),
    ETS = ETS(items_sold),
    SNAIVE = SNAIVE(items_sold),
    TSLM = TSLM(items_sold ~ trend() + season() + weekday + is_holiday + snap_TX),
    PROPHET = prophet(items_sold ~ weekday +  snap_TX)
  )


forecasts_sampled <- models_sampled %>%
  forecast(new_data = future_tsibble %>%
             filter(id %in% sample_ids))


```

```{r}
# Join forecasts with validation data
results_comparison <- forecasts_sampled %>%
  as_tibble() %>%
  select(.model, id, date, forecast = .mean) %>%
  left_join(
    validation_tsibble,  # Validation data containing actual sales
    by = c("id", "date")
  )

# Calculate error metrics for each model and product
results_comparison <- results_comparison %>%
  mutate(
    error = abs(forecast - items_sold)  
  )

# MAE and RMSE 
metrics <- results_comparison %>%
  group_by(.model, id) %>%
  summarise(
    MAE = mean(error, na.rm = TRUE),
    RMSE = sqrt(mean(error^2, na.rm = TRUE)),
    .groups = 'drop' 
  )

print(metrics)
```

```{r}
# Aggregate metrics by model across all products
aggregate_metrics <- metrics %>%
  group_by(.model) %>%
  summarise(
    MAE = mean(MAE, na.rm = TRUE),
    RMSE = mean(RMSE, na.rm = TRUE),
    .groups = 'drop'
  )

# Print aggregate metrics
print(aggregate_metrics)
```

# Final Model

Since ARIMA2 model scored best it will be applied to the whole data set.

```{r}

all_ids <- df_merged_tsibble %>%
  distinct(id) %>%
  pull(id)


future_dates <- tibble(
  date = seq.Date(
    from = max(df_merged_tsibble$date) + 1,
    by = "day",
    length.out = 28
  )
) %>%
  mutate(
    weekday = factor(weekdays(date), levels = c("Sunday", "Monday", "Tuesday", 
                                                "Wednesday", "Thursday", "Friday", 
                                                "Saturday"))
  )

# Add the sampled IDs to align with the model's key
future_tsibble <- expand_grid(
  date = future_dates$date,
  id = all_ids  # Include the sampled product IDs here
) %>%
  left_join(future_dates, by = "date") %>%  # Add the weekday column to the grid
  as_tsibble(index = date, key = id)

future_tsibble <- future_tsibble %>%
  left_join(calendar %>% select(date, month, is_holiday, snap_TX), by = "date")

future_tsibble <- future_tsibble %>%
  left_join(data_sell_daily, by = c("date", "id" = "item_id"))

future_tsibble 
```

```{r}
final_model <- df_merged_tsibble %>%
  model(
    ARIMA(items_sold ~ weekday + snap_TX),
  )


forecasts_final <- final_model %>%
  forecast(new_data = future_tsibble %>%
             filter(id %in% all_ids))

forecasts_final
```

```{r}
# normalise forecast to int 
forecasts_final <- forecasts_final %>%
  mutate(
    .mean = pmax(0, round(.mean))
  )

# Join forecasts with validation data
results_comparison_final <- forecasts_final %>%
  as_tibble() %>%
  select(.model, id, date, forecast = .mean) %>%
  left_join(
    validation_tsibble,  
    by = c("id", "date")
  )

```

```{r}
# Calculate error metrics for each model and product
results_comparison_final <- results_comparison_final %>%
  mutate(
    error = abs(forecast - items_sold) 
  )

# MAE and RMSE 
metrics <- results_comparison_final %>%
  group_by(.model, id) %>%
  summarise(
    MAE = mean(error, na.rm = TRUE),
    RMSE = sqrt(mean(error^2, na.rm = TRUE)),
    .groups = 'drop'  # Prevent grouped data frame issues
  )

print(metrics)
```

```{r}
# Aggregate metrics by model across all products
aggregate_metrics <- metrics %>%
  group_by(.model) %>%
  summarise(
    MAE = mean(MAE, na.rm = TRUE),
    RMSE = mean(RMSE, na.rm = TRUE),
    .groups = 'drop'
  )

# Print aggregate metrics
print(aggregate_metrics)
```

```{r}
# Aggregate residuals
residuals_aggregate <- final_model %>%
  residuals() %>%
  as_tibble() %>%
  group_by(date) %>%
  summarise(
    mean_residual = mean(.resid, na.rm = TRUE),
    sd_residual = sd(.resid, na.rm = TRUE),
    .groups = 'drop'
  )

# Visualize aggregate residuals over time
ggplot(residuals_aggregate, aes(x = date, y = mean_residual)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Mean Residuals Over Time", x = "Date", y = "Mean Residual")

```

```{r}
# convert forecast in rigth sumbissimon format
# Transform `forecasts_final` into wide format

# Add a column cycling from F1 to F28
forecasts_submission <- forecasts_final %>%
  as_tibble() %>%  # Convert the fable to a tibble
  select(id, date, .mean) %>%  # Select relevant columns
  group_by(id) %>%  # Group by `id` to restart the cycle for each product
  mutate(F_day = paste0("F", rep(1:28, length.out = n()))) %>%
  ungroup()  # Ungroup after mutation

# View the result
print(forecasts_submission)
```

```{r}
# Pivot the tibble into the desired format
forecasts_wide <- forecasts_submission %>%
  select(id, F_day, .mean) %>%  # Keep only relevant columns
  pivot_wider(names_from = F_day, values_from = .mean)

# View the resulting tibble
print(forecasts_wide)

# Save the forecasts_wide data frame as a CSV file
write.csv(forecasts_wide, "final_submission.csv", row.names = FALSE)

```
